---
title: "Activity 14"
subtitle: "STAT 184"
author: "Alfred Liljas"
date: "Tuesday, November 11, 2025"
date-format: full
date-modified: now
format:
  html:
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: right
    fig-align: center
    cap-location: top
    link-external-newwindow: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  eval: true
  warning: false
engine: knitr
---

## Introduction

This document contains the R code for activities 4, 8, 10, and 13 regarding armed forced data wrangling, popularity of baby names, and plotting mathematical functions.

## Armed Forces Data Wrangling Redux (Activities #08 and #10)

```{r}
#| echo: false
library(tidyverse)
library(rvest)
library(googlesheets4)

rank_page <- "https://neilhatfield.github.io/Stat184_PayGradeRanks.html"

ranks_raw <- rank_page %>%
  read_html() %>%
  html_elements("table") %>%
  html_table() %>%
  pluck(1)
#clean rank data
ranks_raw[1, 1] <- "Type"
colnames(ranks_raw) <- ranks_raw[1, ]
ranks_trimmed <- ranks_raw[-c(1, 26), ]

ranks_tidy <- ranks_trimmed %>%
  select(-Type) %>%
  pivot_longer(
    cols = -`Pay Grade`,
    names_to = "Branch",
    values_to = "Rank"
  ) %>%
  mutate(Rank = na_if(Rank, "--"))

gs4_deauth()
#read in the armed forces data from Google docs
sheet_url <- "https://docs.google.com/spreadsheets/d/19xQnI1cBh6Jkw7eP8YQuuicMlVDF7Gr-nXCb5qbwb_E/edit?usp=sharing"

headers_preview <- read_sheet(sheet_url, col_names = FALSE, n_max = 3)
#read the data while skipping headers
forces_raw <- read_sheet(
  sheet_url,
  col_names = FALSE,
  skip = 3,
  n_max = 28,
  col_types = "c"
)
#Cleaning the forces data
branches <- c("Army", "Navy", "Marine Corps", "Air Force", "Space Force", "Total")
expanded_branches <- rep(branches, each = 3)
col_names <- paste0(c("", expanded_branches), ".", headers_preview[3, ])

names(forces_raw) <- col_names
#Making the data tidy
forces_tidy <- forces_raw %>%
  rename(Pay.Grade = `.Pay Grade`) %>%
  select(-contains("Total")) %>%
  filter(!Pay.Grade %in% c("Total Enlisted", "Total Warrant Officers", "Total Officers", "Total")) %>%
  pivot_longer(
    cols = -Pay.Grade,
    names_to = "Branch_Sex",
    values_to = "Frequency"
  ) %>%
  separate(Branch_Sex, into = c("Branch", "Sex"), sep = "\\.") %>%
  mutate(
    Frequency = na_if(Frequency, "N/A*"),
    Frequency = parse_number(Frequency)
  )
#merge rank data and forces data
forces_with_ranks <- forces_tidy %>%
  left_join(ranks_tidy, by = c("Pay.Grade" = "Pay Grade", "Branch" = "Branch"))
#create data frame with individual data
soldiers_individual <- forces_with_ranks %>%
  drop_na(Frequency) %>%
  uncount(weights = Frequency)
head(soldiers_individual) %>%
  knitr::kable(caption = "Preview of Individual Soldier Data") %>%
  kableExtra::kable_styling(full_width = FALSE)


```



```{r}
#| echo: false
library(janitor)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)

# Reads in data, adjust file path if needed
military_data <- read.csv("C:\\Users\\alfre\\Downloads\\US Armed Forces (6_2024) - Sheet1.csv", 
                          skip = 1, header = TRUE)

# Rename columns
colnames(military_data) <- c(
  "Pay_Grade", 
  "Army_Male", "Army_Female", "Army_Total",
  "Navy_Male", "Navy_Female", "Navy_Total",
  "Marine_Male", "Marine_Female", "Marine_Total",
  "AirForce_Male", "AirForce_Female", "AirForce_Total",
  "SpaceForce_Male", "SpaceForce_Female", "SpaceForce_Total",
  "Total_Male", "Total_Female", "Total"
)

# Clean data
military_data <- military_data %>%
  filter(!grepl("Source|Pay Grade|Total", Pay_Grade))

# Convert columns to be numeric
military_data[, -1] <- lapply(military_data[, -1], as.numeric)

# Choosing which branch
chosen_branch <- "Army"
branch_data <- military_data %>%
  select(Pay_Grade,
         Male = paste0(chosen_branch, "_Male"),
         Female = paste0(chosen_branch, "_Female"))

branch_data <- branch_data %>%
  mutate(Rank_Category = case_when(
    grepl("^E", Pay_Grade) ~ "Enlisted",
    grepl("^W", Pay_Grade) ~ "Warrant Officer",
    grepl("^O", Pay_Grade) ~ "Officer",
    TRUE ~ "Other"
  ))

# Filter only Warrant Officers
branch_data <- branch_data %>%
  filter(Rank_Category == "Warrant Officer")

# Sum up counts for each gender
freq_table <- branch_data %>%
  summarise(
    Male = sum(Male, na.rm = TRUE),
    Female = sum(Female, na.rm = TRUE)
  ) %>%
  adorn_totals("row")

# Frequency table
freq_table %>%
  kbl(caption = paste("Table 1. Total Number of", chosen_branch, "Warrant Officers by Sex")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Popularity of Baby Names(Activity #13)

I picked the names Emma, Noah, Olivia and Ethan since they were some of the most popular names I could think of. The visualization displays and compares the popularity of female and male names from 1880 to 2020. We can see that all 4 names became much more popular in the last 50 years.

```{r}
library(babynames)
library(dplyr)

names <-c("Emma","Noah", "Olivia", "Ethan")
#Filters out names that aren't the selected ones
babynames_filtered <- babynames %>%
  filter(name %in% names) %>%
  group_by(year,name,sex) %>%
  summarise(total=sum(n), .groups='drop')

ggplot(babynames_filtered, aes(x = year, y = total, color = name, linetype = sex)) +
  geom_line(size=1) +
  labs(
    title = "Popularity of Baby Names Over Time",
    x = "Year",
    y = "Total",
    color = "Name",
    linetype = "Sex"
  ) +
  #Creates two separate graphs
  facet_wrap(~sex) +
  theme(
    plot.title=element_text(face="bold", size=14),
    legend.position="top"
  )

```

## Plotting a Mathematical Function(Activity #4)

This plot shows how the total volume of an open-top box changes as the side length of the cutout increases from 0 to 22 inches. The volume increases sharply until it hits it's peak at about 7 inches, and then it starts decreasing. The plot shows how optimizing the cut size leads to the largest possible volume for a 36 by 48 inch sheet

```{r}
library(ggplot2)

volume <- function(x, length = 36, width = 48) {
  (length - 2*x) * (width - 2*x) * x
}

ggplot(data.frame(x = c(0, 22)), aes(x = x)) +
  stat_function(fun = volume, color = "blue", linewidth = 1.2) +
  labs(
    title = "Box Volume vs. Side Length of Cutout",
    x = "Side Length of Cutout (inches)",
    y = "Volume (cubic inches)"
  ) +
  theme_minimal(base_size = 14)
```

## Reflection

So far in this course, I have learned a lot about the R programming language. It started off with simple plots based on equations like the one from activity #4. Then, I learned how to use packages from R and use datasets found in those packages to create more complicated data visualizations using more complex functions of R. I also learned how to create visualizations with data found on the internet or from files downloaded from my computer. Finally, I have now learnt how to create Quarto markdown files from doing this project. Overall, this class has been a great learning experience with the R programming language.

## Code Appendix

# Activity #8:

```{r}
#| echo: true
#| eval: false
library(tidyverse)
library(rvest)
library(googlesheets4)

rank_page <- "https://neilhatfield.github.io/Stat184_PayGradeRanks.html"

ranks_raw <- rank_page %>%
  read_html() %>%
  html_elements("table") %>%
  html_table() %>%
  pluck(1)
#clean rank data
ranks_raw[1, 1] <- "Type"
colnames(ranks_raw) <- ranks_raw[1, ]
ranks_trimmed <- ranks_raw[-c(1, 26), ]

ranks_tidy <- ranks_trimmed %>%
  select(-Type) %>%
  pivot_longer(
    cols = -`Pay Grade`,
    names_to = "Branch",
    values_to = "Rank"
  ) %>%
  mutate(Rank = na_if(Rank, "--"))

gs4_deauth()
#read in the armed forces data from Google docs
sheet_url <- "https://docs.google.com/spreadsheets/d/19xQnI1cBh6Jkw7eP8YQuuicMlVDF7Gr-nXCb5qbwb_E/edit?usp=sharing"

headers_preview <- read_sheet(sheet_url, col_names = FALSE, n_max = 3)
#read the data while skipping headers
forces_raw <- read_sheet(
  sheet_url,
  col_names = FALSE,
  skip = 3,
  n_max = 28,
  col_types = "c"
)
#Cleaning the forces data
branches <- c("Army", "Navy", "Marine Corps", "Air Force", "Space Force", "Total")
expanded_branches <- rep(branches, each = 3)
col_names <- paste0(c("", expanded_branches), ".", headers_preview[3, ])

names(forces_raw) <- col_names

forces_tidy <- forces_raw %>%
  rename(Pay.Grade = `.Pay Grade`) %>%
  select(-contains("Total")) %>%
  filter(!Pay.Grade %in% c("Total Enlisted", "Total Warrant Officers", "Total Officers", "Total")) %>%
  pivot_longer(
    cols = -Pay.Grade,
    names_to = "Branch_Sex",
    values_to = "Frequency"
  ) %>%
  separate(Branch_Sex, into = c("Branch", "Sex"), sep = "\\.") %>%
  mutate(
    Frequency = na_if(Frequency, "N/A*"),
    Frequency = parse_number(Frequency)
  )
#merge rank data and forces data
forces_with_ranks <- forces_tidy %>%
  left_join(ranks_tidy, by = c("Pay.Grade" = "Pay Grade", "Branch" = "Branch"))
#create data frame with individual data
soldiers_individual <- forces_with_ranks %>%
  drop_na(Frequency) %>%
  uncount(weights = Frequency)
head(soldiers_individual) %>%
  knitr::kable(caption = "Preview of Individual Soldier Data") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

# Activity #10:

```{r}
#| echo: true
#| eval: false
library(janitor)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)

# Reads in data, adjust file path if needed
military_data <- read.csv("C:\\Users\\alfre\\Downloads\\US Armed Forces (6_2024) - Sheet1.csv", 
                          skip = 1, header = TRUE)

# Rename columns
colnames(military_data) <- c(
  "Pay_Grade", 
  "Army_Male", "Army_Female", "Army_Total",
  "Navy_Male", "Navy_Female", "Navy_Total",
  "Marine_Male", "Marine_Female", "Marine_Total",
  "AirForce_Male", "AirForce_Female", "AirForce_Total",
  "SpaceForce_Male", "SpaceForce_Female", "SpaceForce_Total",
  "Total_Male", "Total_Female", "Total"
)

# Clean data
military_data <- military_data %>%
  filter(!grepl("Source|Pay Grade|Total", Pay_Grade))

# Convert columns to be numeric
military_data[, -1] <- lapply(military_data[, -1], as.numeric)

# Choosing which branch
chosen_branch <- "Army"
branch_data <- military_data %>%
  select(Pay_Grade,
         Male = paste0(chosen_branch, "_Male"),
         Female = paste0(chosen_branch, "_Female"))

branch_data <- branch_data %>%
  mutate(Rank_Category = case_when(
    grepl("^E", Pay_Grade) ~ "Enlisted",
    grepl("^W", Pay_Grade) ~ "Warrant Officer",
    grepl("^O", Pay_Grade) ~ "Officer",
    TRUE ~ "Other"
  ))

# Filter only Warrant Officers
branch_data <- branch_data %>%
  filter(Rank_Category == "Warrant Officer")

# Sum up counts for each gender
freq_table <- branch_data %>%
  summarise(
    Male = sum(Male, na.rm = TRUE),
    Female = sum(Female, na.rm = TRUE)
  ) %>%
  adorn_totals("row")

# Frequency table
freq_table %>%
  kbl(caption = paste("Table 1. Total Number of", chosen_branch, "Warrant Officers by Sex")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

# Activity #13:

```{r}
#| echo: true
#| eval: false
#| alt: "Line graph of baby name popularity over time for Emma, Olivia (female) and Ethan, Noah (male), from 1880 to 2020."
# ALT TEXT: Line graph of baby name popularity over time for Emma, Olivia (female) and Ethan, Noah (male), from 1880 to 2020.

library(babynames)
library(dplyr)

names <-c("Emma","Noah", "Olivia", "Ethan")
#Filters out names that aren't the selected ones
babynames_filtered <- babynames %>%
  filter(name %in% names) %>%
  group_by(year,name,sex) %>%
  summarise(total=sum(n), .groups='drop')

ggplot(babynames_filtered, aes(x = year, y = total, color = name, linetype = sex)) +
  geom_line(size=1) +
  labs(
    title = "Popularity of Baby Names Over Time",
    x = "Year",
    y = "Total",
    color = "Name",
    linetype = "Sex"
  ) +
  #Creates two separate graphs
  facet_wrap(~sex) +
  theme(
    plot.title=element_text(face="bold", size=14),
    legend.position="top"
  )
```

# Activity #4:

```{r}
#| echo: true
#| eval: false
#| alt: "Graph showing Box Volume vs. Side Length of Cutout with a blue curve that peaks near 8 inches."
#ALT TEXT: Graph showing Box Volume vs. Side Length of Cutout with a blue curve that peaks near 8 inches.

library(ggplot2)

volume <- function(x, length = 36, width = 48) {
  (length - 2*x) * (width - 2*x) * x
}

ggplot(data.frame(x = c(0, 22)), aes(x = x)) +
  stat_function(fun = volume, color = "blue", linewidth = 1.2) +
  labs(
    title = "Box Volume vs. Cut Size",
    x = "Side Length of Cutout (inches)",
    y = "Volume (cubic inches)"
  ) +
  theme_minimal(base_size = 14)
```
